### 2014-06-02 text/Chapter04.tex
Syntactic properties (Chapter 4)

\subsection{Call-by-value evaluation}
\subsection{Call-by-name evaluation}

@@

Operational semantics of every context-dependent language differs as the notion of context 
is always different. However, for coeffect calculi satisfying certain conditions we can 
define a universal equational theory. This suggests a pathway to an operational semantics for 
two out of our three examples (the notion of context for data-flow is more complex).

In a pure $\lambda$-calculus, $\beta$ and $\eta$ equality for
functions (also called \emph{local soundness} and \emph{completeness}
respectively~\cite{logic-modal-reconstruction})
describe how pairs of abstraction and application can be eliminated:
$(\lambda x . e_2) e_1 \equiv_\beta \subst{e_1}{x}{e_2}$ and $(\lambda x . e \, x) 
\equiv_\eta e$. The $\beta$ equality rule, using the usual Barendregt convention of 
syntactic substitution, implies a \emph{reduction},
giving part of an operational semantics for the calculus.

The call-by-name evaluation strategy modelled by $\beta$-reduction is
not suitable for impure calculi therefore a restricted $\beta$ rule,
corresponding to call-by-value, is used, \ie~$(\lambda x . e_2) v
\equiv \subst{e_2}{x}{v}$. Such reduction can be encoded by a
\emph{let}-binding term, $\kvd{let}~x=e_1~\kvd{in}~e_2$, which
corresponds to sequential composition of two computations, where the
resulting pure value of $e_1$ is substituted into
$e_2$~\cite{monads-inaction,monad-notions}.

%
% We consider here both a notion of \emph{let}-binding for the coeffect
% calculus, useful for a CBV evaluation, and a notion of substitution 
% for a CBV evaluation.
%

>>>>

Each of the concrete coeffect calculi discussed in this chapter has a different notion of context,
much like various effectful languages have different notions of effects (such as exceptions or
mutable state). However, in all of the calculi, the context has a number of common properties that 
are captured by the \emph{flat coeffect algebra}. This means that there are equational properties
that hold for all of the systems we consider. Further properties hold for systems where the context
satisfies additional properties.

In this section, we look at such shared syntactic properties. This accompanies the previous section, 
which provided a \emph{semantic} justification for the axioms of coeffect algebra with a 
\emph{syntactic} justification. Operationally, this section can also be viewed as providing a 
pathway to an operational semantics for two of our systems (implicit parameters and liveness),
which can be based on syntactic substitution. As we discuss later, the notion of context for
data-flow is more complex.

\subsection{Syntactic properties}

Before discussing the syntactic properties of general coeffect calculus formally, it should be 
clarified what is meant by providing ``pathway to operational semantics'' in this section. We do
that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
Assuming $\subst{e_1}{x}{e_2}$ is a standard capture-avoiding syntactic substitution, the following
equations define four syntactic reductions on the terms:
%
\begin{equation*}
\begin{array}{rllcl}
(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized substitution})\\
e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
\end{array}
\end{equation*}
%
The rules capture syntactic reductions that can be performed in a general calculus, without any 
knowledge of the specific notion of context. The \kvd{glet} notation models explicit sequencing
of context-dependent computations an is inspired by Filinski \cite{monads-inaction}. In the rest
of the section, we briefly outline the interpretation of the four rules and then we focus on
call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name (Section~\ref{sec:flat-syntax-cbn})
in more details.

The focus of this work is on the general coeffect system and so we do not discuss the operational
semantics of the specific notions of context. However, some work in that area has been done
by Brunel et al. \cite{coeffects-quantitative}.

\paragraph{Call-by-name.} 
In call-by-name, the argument is syntactically substituted for all occurrences of a variable. It
can be used as the basis for operational semantics of purely functional languages. However, using
the rule in effectful languages breaks the \emph{type preservation} property. For example, consider
a language with effect system where functions are annotated with sets of effects such as $\{ \ident{write} \}$.
A function $\lambda x.y$ is a effect-free:
%
\begin{equation*}
y\!:\!\tau_1 \vdash \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 \;\&\; \emptyset
\end{equation*}
%
Substituting an expression $e$ with effects $\{ \ident{write} \}$ for $y$ changes the type of 
the function by adding latent effects (without changing the immediate effects):
%
\begin{equation*}
\vdash \lambda x.e : \tau_1 \xrightarrow{ \{ \ident{write} \} } \tau_2 \;\&\; \emptyset
\end{equation*}
%
Similarly to effect systems, substituting a context-dependent computation $e$ for a variable $y$ can
add latent coeffects to the function type. However, this is not the case for \emph{all} flat coeffect
calculi. For example, call-by-name reduction preserves types and coeffects for the implicit 
parameters system. This makes the model suitable for languages such as Haskell.

\paragraph{Call-by-value.}

The call-by-value evaluation strategy is often used by effectful languages. Here, an argument is
first reduced to a \emph{value} before performing the substitution. In effectful languages, 
value is defined syntactically. For example, in the \emph{Effect} language \cite{monads-effects-marriage},
values are identifiers $x$ or functions $(\lambda x.e)$.

The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
A function $(\lambda x.e)$ does not delay all context requirements of the body $e$ and may have
immediate context requirements. Thus we say that $e$ is a value if it is a value in the usual
sense \emph{and} has not immediate context requirements. We define this formally in 
Section~\ref{sec:flat-syntax-cbv}.

The call-by-value evaluation strategy holds for a wide range of flat coeffect calculi, including all
our three examples. However, it is rather weak -- in order to use it, the concrete semantics needs
to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ to
a term $\coctx{\Gamma}{\cunit} e' : \tau$ with no context requirements.

\paragraph{Internalized substitution.}

The (\emph{internalized substitution}) rule models an operational semantics where the language
provides a construct that models \emph{sequential composition} and expressions can be reduced to 
a normal form, consisting of a sequenced context-dependent operations. We choose to write the 
sequencing operation as \kvd{glet} to emphasize that this is a separate primitive and not an 
ordinary syntactic \kvd{let}. The normal form looks as follows:
%
\begin{equation*}
\kvd{glet}~x_1 = e_1~\kvd{in}~\ldots~\kvd{glet}~x_n = e_n~\kvd{in}~e
\end{equation*}
%
Here, the expressions $e_1, \ldots, e_n, e$ do not contain further nested \kvd{glet} constructs.
This evaluation strategy provides context-dependent counterpart to operational view of monads
developed by Filinsky \cite{monads-inaction}. We discuss how expressions reduce to the normal form
in Section~\ref{sec:flat-syntax-norm}

The (\emph{internalized substitution}) rule is useful when defining a concrete semantics for 
a language that provides constructs for explicitly providing the context. For example, consider
a language with implicit parameters where a parameter is defined by $e_1~\kvd{with}~\ident{?p}=e_2$.
Semantics for such language would provide a reduction rule:
%
\begin{equation*}
(\kvd{glet}~x_1 = \ident{?p}~\kvd{in}~e)~\kvd{with}~\ident{?p}=e_1 ~~\leadsto~~ \kvd{let}~x_1 = e_1~\kvd{in}~e
\end{equation*}
%
Here, the \kvd{glet} construct provides a way of sequentialising the context-requi\-rements so
that they can be discharged by matching constructs providing the required contexts. As discussed
earlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
this form of evaluation (rather than looking at semantics for concrete systems).

\paragraph{Local soundness and completeness.}
Two desirable properties of calculi, coined by Pfenning and Davies \cite{logic-modal-reconstruction},
are \emph{local soundness} and \emph{local completeness}. They guarantee that the rules which 
introduce a function arrow (lambda abstraction) and eliminate it (application) are not too strong
and sufficiently strong.

The local soundness property is witnessed by (call-by-name) $\beta$-reduction, which we discussed
already. The local completeness is witnessed by the $\eta$-expansion rule. We discuss the flat
coeffect algebra conditions under which the reduction holds in Section~\ref{sec:flat-syntax-cbn}.

\subsection{Call-by-value evaluation}
\label{sec:flat-syntax-cbv}

\begin{definition} 
An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if $\coctx{\Gamma}{ \aclrd{\mathsf{VAL}} } \vdash e : \tau$ 
where $\aclrd{\mathsf{VAL}} : \C^{mksh{\Gamma}}$ is a coeffect indexed by the shape of $\Gamma$ that
always returns $\cunit$. That is $\aclrd{\mathsf{VAL}} = \lambda \sclrd{n} . \cunit$.
\end{definition}

Recall that a flat coeffect system consists of coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
together with a binary operation $\cclrd{\wedge}$ on $\C$. The coeffect algebra is then defined
as $(\cclrd{\wedge}, \cpar, \czero)$ where $\cclrd{\wedge}$ and $\cpar$ represent splitting and
merging of context-requirements, respectively.

\begin{lemma}[Call-by-value substitution]
\label{thm:flat-subst-cbv}
In a flat coeffect calculus with coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
and the $\cclrd{\wedge}$ operator:
\[
\begin{array}{l}
\coctx{\Gamma}{\aclrd{\mathsf{VAL}}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\end{lemma}
\begin{proof}
By induction over the type derivation, using the fact that both $x$ and $e_s$ are annotated
with $\cunit$.
\end{proof}

\noindent
Lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems, but it is weak. To use it, 
the operational semantics must provide a way of partially evaluating a term 
with requirements $\coctx{\Gamma}{\cclrd{r}}$ to a value.
Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction using the above definition of value:

\begin{theorem}[Call-by-value reduction]
In a flat coeffect system, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
A direct consequence of Lemma~\ref{thm:flat-subst-cbv}, using the flat coeffect system requirement
$(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$ to prove $\beta$-equality.
\end{proof}

\subsection{Call-by-name evaluation}
\label{sec:flat-syntax-cbn}

The call-by-name strategy reduces a term $(\lambda x. e_1)~e_2$ where both sub-expressions 
may have contextual requirements. Assume that $\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2$ and
$\coctx{\Gamma_1, x :\tau_1}{\cclrd{r}} \vdash e_1 : \tau_1$.

% TODO
%
% The context requirements of the 
% reduced term $\subst{e_1}{x}{e_2}$ should be the same as the context requirements of 
% $\kvd{let}~x=e_2~\kvd{in}~e_1$. Thus the context requirements of the result in the substitution 
% lemma should be $r \fmax (s \fseq r)$.

We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
$\C$ and \emph{bottom-pointed} if it is the smallest (bottom) coeffect scalar.
Liveness analysis is an example of top-pointed coeffects
as variables are annotated with $\ident{L}$ and $\ident{D} \leq \ident{L}$.

\begin{lemma}[Top-pointed substitution]
\label{thm:cbn-substitution-top}
In a top-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
and the $\cclrd{\wedge}$ operator:
\[
\begin{array}{l}
\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\end{lemma}
\begin{proof}
Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and Lemma~\ref{thm:flat-subst-cbv}.
\end{proof}
%
\noindent
As variables are annotated with the top element $\cunit$, we can substitute a term $e_s$ for any variable and 
use sub-coeffecting to get the original typing (because $\cclrd{s} \,\cleq\, \cunit$).

In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
requirements. However, if the system satisfies the strong condition that $\cclrd{\wedge} = \cseq = \cpar$ then
the context requirements arising from the substitution can be associated with the context
$\Gamma$. As a result, substitution does not break soundness as in effect systems. The 
requirement $\cclrd{\wedge} = \cseq = \cpar$ holds for our implicit parameters example 
(all three operators are set union) and allows the following substitution lemma:

\begin{lemma}[Bottom-pointed substitution]
\label{thm:cbn-substitution-bot}
In a bottom-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
and the $\cclrd{\wedge}$ operator where $\cclrd{\wedge} = \cseq = \cpar$ is idempotent and commutative:
\[
\begin{array}{l}
\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\end{lemma}
\begin{proof}
By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
the free-variable context.
\end{proof}
\noindent
The structural system is precise enough to keep the coeffects associated with a concrete variable. The flat
variant described here is flexible enough to let us always re-associate new context requirements with the 
free-variable context.

The two substitution lemmas show that the call-by-name evaluation strategy can be used for
certain coeffect calculi, including liveness and implicit parameters.
Assuming $\reduce_{\ident{cbn}}$ is the standard call-by-name
reduction, the following theorem holds:

\begin{theorem}[Call-by-name reduction]
\raggedright
In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
and $e \rightarrow_{\ident{cbn}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
A direct consequence of Lemma~\ref{thm:cbn-substitution-top} or Lemma~\ref{thm:cbn-substitution-bot}.
\end{proof}

ALSO: Something on eta expansion

\subsection{Normal form}
\label{sec:flat-syntax-norm}

\begin{equation*}
\tyrule{glet}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_2 }
  { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}

First transformation
\begin{equation*}
(\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3  ~~\leadsto~~ \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
\end{equation*}

Second transformation

\begin{equation*}
\kvd{glet}~x_2=(\kvd{glet}~x_1 = e_1~\kvd{in}~e_2)~\kvd{in}~e_3  ~~\leadsto~~
\kvd{glet}~x_1=e_1~\kvd{in}~(\kvd{glet}~x_2=e_2~\kvd{in}~e_3)
\end{equation*}

or

%
% TODO: We could have a section about let binding covering the alternative CBV and the let binding
% we included in the system, but we'd have to clean this up!
%
%\subsection{Syntactic let binding}
%\begin{remark}
%As mentioned earlier, the bounded reuse coeffect system in Section~\ref{}
%is for a CBN semantics. Recall the coeffect judgment for \emph{let} in the bounded reuse
%calculus of Section~\ref{}.
%%%
%\begin{align*}
%{\footnotesize{\trule{let}}}
%\dfrac
%{\Gamma ? \B{S} \vdash e_1 : \tau_1 \quad
%\Gamma, v : \tau_1 ? \B{R} \times T \vdash e_2 : \tau_2}
%{\Gamma ? \B{R} + (T \ast \B{S}) \vdash \synLet{v}{e_1}{e_2} : \tau_2}
%\end{align*}
%%%
%Therefore, for example, we see that \emph{let} has the coeffects
%as substitution, for example:
%%%
%\begin{align*}
%& {\footnotesize{\trule{let}}}
%\dfrac
%{y : \tau ? \langle{2}\rangle \vdash y + y : \tau \quad
%y : \tau, v : \tau ? \langle{1, 2}\rangle \vdash y + v + v : \tau_2}
%{\Gamma ? \langle{5}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2} \\
%& \hspace{3.55em} \equiv \; \Gamma ? \langle{5}\rangle \vdash y + (y + y) + (y + y) : \tau_2
%\end{align*}
%In a CBV version of bounded reuse however, the \emph{CBV let} rule gives us instead
%the judgment:
%%%
%$$
%\Gamma ? \langle{3}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2
%$$
%%%
%\end{remark}
%
% 
%\paragraph{Syntactical let binding}
%Has the same typing derivation as abstraction/application
%
%\[
%\inference
% {\inference
%    { \coctx{\Gamma_1, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{t}}} \vdash e_1 : \tau }
%    { \coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{t}} \tau }
%  &
%    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
%  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash (\lambda x. e_1)~e_2 : \tau } 
%\]
%
%We can extend the CBV behaviour to \emph{let}, where the bound term
%is evaluated to a pure value first, discharging any contextual
%requirements before substitution into the body.
%
%\begin{lemma}[CBV \emph{let}]
%Let $\Gamma, x : \tau' ? \B{R} \times T \vdash e : \tau$ and 
%$\Gamma ? \B{S} \vdash e' : \tau'$. Then $\Gamma ? \B{R} \sqcup \B{S} \vdash \synLet{x}{e'}{e}
%: \tau$.
%\end{lemma}

@@
